\documentclass[11pt]{article}
\ifdefined\XeTeXrevision\else\pdfoutput=1\fi

\usepackage[margin=1.1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,bm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{float}
\usepackage{microtype}
\usepackage{caption}
\captionsetup{font=small,skip=8pt,justification=raggedright,singlelinecheck=false}
\captionsetup[table]{position=above}
\input{macros}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\hypersetup{
  pdftitle={A Differentiable, Exact Selected-Inverse Layer for PyTorch:
            Size Extrapolation and Stiff-Regime Trainability, Explained by
            Attainability, Not Inductive Bias},
  pdfauthor={Vanshdeep Sehrawat},
}

\setlength{\emergencystretch}{1em}

\raggedbottom

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

\begin{document}

\begin{titlepage}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\begin{center}
\vspace*{1.2cm}

{\LARGE A Differentiable, Exact Selected-Inverse Layer for PyTorch:\\[0.45em]
Size Extrapolation and Stiff-Regime Trainability,\\[0.25em]
Explained by Attainability, Not Inductive Bias\par}

\vspace{1.6cm}

{\large Vanshdeep Sehrawat\footnote{e-mail:
  \texttt{sehrawat@unlv.nevada.edu}, \texttt{vanshdeep.sehrawat@gmail.com};\ ORCID:
  \href{https://orcid.org/0009-0005-8062-1337}{0009-0005-8062-1337}.}}\\[0.35em]
{\itshape University of Nevada, Las Vegas, USA}

\vspace{0.9em}

{\small Code, data, and frozen pre-registration records:
\href{https://github.com/vanshsehrawat14/gabp-sparse-inv}{\texttt{github.com/vanshsehrawat14/gabp-sparse-inv}}
(package \texttt{gabp-sparse-inv} on PyPI).\\
File paths cited in this paper are relative to the repository root.}

\vspace{1.2cm}

{\bfseries Abstract}
\end{center}

\begin{quote}
Any layer built on self-consistency, soft constraint satisfaction, or a fixed
point pays for a matrix inverse; deep equilibrium models, delta-rule linear
attention, and implicit graph layers are all of this form. When the coupling
is sparse on a low-treewidth graph, the \emph{selected} inverse (the blocks of
$\inv{A}$ on the sparsity pattern of $A$) is exact in $\bigO(n)$ time by
Gaussian belief propagation and the Takahashi recurrences, and its
reverse-mode adjoint is again a selected inversion at the same cost. This is
implemented as differentiable PyTorch operators, symmetric and non-symmetric.
The operator is classical and buys no precision over a well-ordered dense
solve (Section~\ref{sec:numerics}). The hypothesis under test was that
embedding it buys something else: that \emph{exactness is an inductive
bias} -- that among models reaching comparable training fit, the one
containing the exact global solve is preferred and generalizes better. Two
pre-registered probes on $30$ held-out seeds test this directly with a
matched-fit design that asks not just whether the exact-trained model
generalizes better, but whether the compared truncated/iterative model can
even \emph{reach} comparable training fit at all. On both probes it cannot:
in the high-difficulty cells the truncated model matches the exact model's
training fit in at most $0$--$23\%$ of seeds against a pre-registered
$60\%$ bar, so the held-out gap is explained by \emph{attainability} -- the
approximation cannot fit the task -- not by an implicit preference among
optima that fit equally well. The inductive-bias reading did not survive on
either probe. This does not change the raw comparison: the exact model still
beats matched-or-greater-capacity GNNs and Transformers by three to four
orders of magnitude at every tested size, and the exact selected-inverse
backward stays at machine precision as a fixed point stiffens
($\rho(J)\to1$) where iterative adjoints are $77$ to $100\%$ wrong. What
changes is the explanation for why. One negative control did not fully pass
its pre-specified clause either, so even the conservative attainability
reading is reported as hedged, not pristine, per the frozen decision rules.
The paper specifies where exactness should and should not help at
benchmark scale and names the one prerequisite, a GPU-batched kernel, that
gates testing there; that program tests numerical properties (exact
gradients for calibrated uncertainty, conditioning-independent adjoints, a
resolvent layer against over-squashing), not inductive bias. The present
evidence is controlled and CPU-scale.
\end{quote}

\vfill
\begin{center}
July 2026
\end{center}
\end{titlepage}

\input{sections/introduction}
\input{sections/problem}
\input{sections/theorem}
\input{sections/kernels}
\input{sections/numerics}
\input{sections/demonstrations}
\input{sections/program}
\input{sections/related}
\input{sections/conclusion}

\section*{Acknowledgements}
I thank Isaac Liao (Carnegie Mellon University) for scoping the
selected-inverse operator, sharpening the target regime, and giving feedback on
the maze demonstration and research direction.

\section*{AI assistance disclosure}
I used AI assistance during software development and manuscript preparation,
including coding support, editing, and artifact checks. I reviewed and validated
the assisted work and remain responsible for the code, data, citations, and
claims in this paper.

\appendix
\input{sections/appendix}

\bibliographystyle{plain}
\bibliography{refs}

\end{document}
