\section{Introduction}
\label{sec:intro}

Anywhere a model relies on self-consistency, soft constraint satisfaction, or a
fixed point, it pays for a matrix inverse. Deep equilibrium models
differentiate through an equilibrium with the implicit function theorem
\cite{bai2019deep}; linear-attention and delta-rule layers mix tokens with a
triangular solve \cite{yang2024parallelizing}; implicit and steady-state graph
layers solve for a fixed point. In each case the forward pass, and almost
always the backward pass, is a linear solve. The general inverse is
$\bigO(n^3)$. When the coupling is sparse on a low-treewidth graph, the
\emph{selected} inverse, meaning only the blocks of $\inv{A}$ on the sparsity
pattern of $A$, is exact in $\bigO(n)$ on trees and $\bigO(n^{1.5})$ on 2-D
grids via Gaussian belief propagation and the Takahashi recurrence
\cite{takahashi1973,erismantinney1975}, never forming the dense inverse. Few
systems expose this operation as a learnable layer, and fewer still expose a
differentiable one.

\paragraph{The wrong question, and the right one.}
The obvious pitch is speed or accuracy: a faster, more exact inverse. This
paper resists it, because it is mostly false. On the precision axis, the one
place an exact method should dominate, the selected inverse is a wash: scored on the
pattern against a 64-bit oracle and swept over conditioning, it matches a
well-ordered dense solve, with no robust advantage
(Section~\ref{sec:numerics}). If exactness mattered only as accuracy, there
would be no paper. The right question is different:
\begin{quote}
\emph{When does the exactness of a structured linear-algebra primitive change
what a neural network can learn or generalize, as opposed to merely how fast
or how accurately it computes a fixed quantity?}
\end{quote}

\paragraph{Hypothesis under test: exactness as an inductive bias.}
The hypothesis was that an exact, scale-free, differentiable global solve is a
distinct inductive bias, and that this bias, not accuracy, is what it buys:
exactness yields no numerical advantage (Section~\ref{sec:numerics}) but was
expected to yield a decisive learning advantage \emph{because} it changes
which of two equally-fit models is preferred (Sections~\ref{sec:maze}
and~\ref{sec:deq}). Two pre-registered probes test this on $30$ held-out
seeds with a matched-fit design (Appendix data in
\texttt{archive/docs/CONFIRMATORY\_RESULTS.md}): they ask not only whether the
exact-trained model generalizes better, but whether the alternative can even
reach comparable \emph{training} fit at all. On both probes the
inductive-bias (equal-fit, different-preference) branch never fires: the
alternative is unmatchable at the difficulty levels where the raw gap is
largest ($0$--$23\%$ of held-out seeds reach a pre-registered $60\%$
matchability bar). The raw comparisons below are real and large; the
explanation the paper tests them against is attainability, not inductive
bias, and that is the conservative reading the pre-registration resolves to
(Section~\ref{sec:program} states the honest scope).
\begin{itemize}
  \item \emph{Size extrapolation.} A model that contains the exact solve,
        with size-independent parameters, stays accurate as the problem
        grows, while GNNs and Transformers of more than $250\times$ its
        capacity, trained on the same data, collapse toward the
        predict-the-mean floor (Section~\ref{sec:maze}). A within-architecture
        control, holding the model fixed and replacing the exact solve with
        $K$ truncated sweeps, shows a monotone dose-response in which only the
        exact, global solve generalizes -- and the matched-fit check shows the
        truncated variants cannot reach the exact arm's training fit at any
        swept $K$, so the effect is better read as an attainability/capacity
        gap (a bounded-reach solve cannot represent a globally-coupled target)
        than as an inductive bias among equally-attainable optima.
  \item \emph{Trainability at the edge of stability.} The exact
        $\bigO(\text{fill})$ implicit-differentiation backward stays at
        machine precision as a fixed point stiffens ($\rho(J)\to 1$), exactly
        where iterative adjoints return gradients that are $77$ to $100\%$
        wrong (Section~\ref{sec:deq}). A pre-registered confirmatory run on
        $30$ held-out seeds, with the forward solve held bit-identical
        between arms, shows the same pattern as size extrapolation: the
        iterative-adjoint arm cannot match the exact arm's training fit in
        the stiff cells, so the held-out gap is attainability, not a
        preference among equally-good optima. One negative control (a
        near-machine-precision low-stiffness cell) did not cleanly pass its
        pre-specified clause, and this is disclosed rather than smoothed
        over.
\end{itemize}

\paragraph{What is, and is not, the contribution.}
The exact forward engine is classical
\cite{takahashi1973,erismantinney1975,lin2011selinv,weiss2001correctness}.
That its differentiable adjoint is again a selected inversion at forward cost
is also not new: selected inversion is the gradient of a log-determinant
\cite{zhu2019sparse,dwyer1948symbolic}, so its reverse mode costs a constant
multiple of the forward \cite{baur1983complexity}. Both are stated once, as
scaffolding (Section~\ref{sec:theorem}). The contribution is the scientific
claim and its evidence:
\begin{enumerate}
  \item the paradox that exactness is not accuracy
        (Section~\ref{sec:numerics});
  \item two pre-registered, matched-fit probes that test exactness as an
        inductive bias, one for size generalization with a
        within-architecture control (Section~\ref{sec:maze}), one for
        trainability at the edge of stability (Section~\ref{sec:deq}) --
        and the honest result that both resolve to attainability, not
        inductive bias, under the frozen decision rules;
  \item a falsifiable benchmark program (Section~\ref{sec:program})
        stating where exactness's numerical properties (not the retired
        inductive-bias claim) are expected to matter, where they must not (a
        negative control), and the single build prerequisite that gates
        testing it at benchmark scale.
\end{enumerate}

\paragraph{Scope.}
The present evidence is at controlled, CPU scale: routing grids up to
$10\times10$ for the baseline comparison and up to $48\times48$ for the
held-out matched-compute control, and deep-equilibrium fixed points at
$n\le16$. The structural win is confined
to low-treewidth (1-D and quasi-2-D) problems; dense 3-D problems belong to
multigrid. The benchmark-scale tests in Section~\ref{sec:program} are gated on
one named, scoped engineering item, a GPU-batched kernel, not on new theory.
