\section{The primitive: an exact, differentiable global sparse solve}
\label{sec:primitive}

The vehicle for the operator family under test is a differentiable selected-inverse layer.
The later sections need three facts about it: it is exact and global
(\S\ref{sec:setup}), it is differentiable at the asymptotic cost of its
forward pass (\S\ref{sec:theorem}), and it is available across the matrix
structures the experiments use (\S\ref{sec:kernels}). None of the three is
new. Together they make exactness embeddable and trainable.

\subsection{Selected inversion on a pattern}
\label{sec:setup}

Let $A \in \R^{N\times N}$ be a block matrix with $n$ diagonal blocks of size
$b$ ($N = nb$) whose block sparsity pattern is a graph $\Gr$ on the $n$ nodes.
The \emph{selected inverse} on a pattern $\pat \supseteq \Gr$ is
\begin{equation}
  \selinv_{\pat}(A) \;=\; \{\, \Gblock{u}{v} \;:\; (u,v)\in\pat \,\},
  \qquad \Gblock{u}{v} = \big(\inv{A}\big)_{uv},
  \label{eq:selinv-def}
\end{equation}
that is, the blocks of $\inv{A}$ on $\pat$ only, never the dense inverse. For
SPD $A$ the natural pattern is the pattern of $A$ itself when $\Gr$ is a tree
(zero fill), and the filled (chordal) pattern
$\pat = \mathrm{pattern}(L+L\T)$ otherwise. Most entries of $\inv{A}$ are dense
and cannot even be materialized at scale; the on-pattern blocks are exactly
what marginal-variance and log-determinant-gradient computations read.  The
same sparse factors also supply the solve siblings used for routed fields
$\inv{A}e_s$ and fixed-point backwards.

\paragraph{Why it is global, exact, and structurally sparse.}
$\inv{A}$ couples every node to every other: $\Gblock{u}{v}$ is a global
functional of the whole graph, not of a local neighbourhood. Yet if $\Gr$ has
low treewidth, a fill-reducing elimination order keeps the elimination fronts
small. More precisely, let $U_v$ be the later neighbours of node $v$ in the
filled graph, and define
\[
  F=\sum_v(1+|U_v|),\qquad W=\sum_v(1+|U_v|^2).
\]
$F$ counts factor-pattern blocks, whereas $W$ counts dense clique work.
Numeric storage is $\Theta(Fb^2)$ and factorization plus selected inversion is
$\Theta(Wb^3)$. They are proportional at bounded treewidth, but not in
general: both are $\Theta(n)$ on trees, while nested dissection on a 2-D grid
has $F=\Theta(n\log n)$ and $W=\Theta(n^{3/2})$. The Takahashi recurrence
computes \eqref{eq:selinv-def} algebraically exactly, up to floating-point
rounding, rather than by a convergence-truncated iteration. Its work depends
on elimination structure, not on the number of message-passing iterations
needed to traverse the graph.

\paragraph{The target regime.}
The target is well-scaled operators with large condition number, e.g.\
graph-Laplacian precisions $A = L(W) + \varepsilon I$ with $\bigO(1)$ entries
and $\kappa \approx \lambda_{\max}/\varepsilon$. SPD Cholesky needs no
pivoting, but its forward accuracy remains condition-sensitive; ``direct''
means iteration-independent, not condition-independent. The non-symmetric
kernel additionally assumes a static no-pivot pattern and nonsingular Schur
pivots. These restrictions are material when the coupling is long-range.
