\section{Related work}
\label{sec:related}

\paragraph{The exact engine.}
The Takahashi equations \cite{takahashi1973} and Erisman--Tinney
\cite{erismantinney1975} compute inverse entries on the Cholesky factor's
pattern without the full inverse; SelInv \cite{lin2011selinv} and PEXSI
\cite{lin2013pexsi} do this for general sparse SPD and Hermitian matrices via
the elimination tree; Gaussian belief propagation is the same
collect/distribute recurrence and is exact on trees
\cite{weiss2001correctness,bickson2008gaussian}. Recent accelerator systems
include \texttt{Serinv} for block-tridiagonal arrowhead matrices
\cite{maillou2025serinv} and tile-parallel selected inversion for structured
matrices \cite{fattah2025stiles}. This work is a reference PyTorch
implementation, not a performance competitor to those systems, and claims
nothing new for the forward recurrence.

\paragraph{Differentiating selected inversion.}
The reverse-mode adjoint of the matrix inverse,
$\cotan{A}=-\inv{A}\cotan{G}\inv{A}$, is folklore
\cite{dwyer1948symbolic,giles2008collected}; that its evaluation costs a
constant factor of the forward is the cheap-gradient principle
\cite{baur1983complexity}; and differentiating a sparse log-determinant
through selected inversion on the same factorization is done by
\cite{zhu2019sparse} and is standard in GMRF and sparse-Cholesky learning
\cite{rue2005gmrf}. TMB has long combined sparse latent-Gaussian calculations,
automatic differentiation, and Laplace approximation at large scale
\cite{kristensen2016tmb}. These results provide the mathematical background for
Remark~\ref{rem:scaffold}.

\paragraph{Differentiable sparse linear algebra.}
Schwarz--Schur Involution batches direct sparse subdomain operations on GPUs
and supplies a differentiable solver for PDE-shaped problems
\cite{wang2025schwarz}. \texttt{torch-sla} unifies differentiable direct,
iterative, nonlinear, and eigenvalue solvers across CPU, GPU, and distributed
backends \cite{torchsla2026}. These systems make a generic
``differentiable sparse solver'' claim untenable. The narrower software
surface here is selected covariance entries, log determinant, solve, and
sampling, including a static-pattern non-symmetric selected inverse. The
paper's scientific contribution is
instead the pre-registered account
(Sections~\ref{sec:maze}--\ref{sec:deq}) of why exactness wins on its
solve-shaped probes: attainability against bounded-reach or biased-gradient
alternatives, not an inductive-bias effect. Approximate selected-covariance
methods with estimable confidence bounds \cite{siden2017covariance} are
important competitors for the prospective uncertainty experiment.

\paragraph{Where the claim lives.}
The original motivation drew on the lineage of inductive bias for
generalization and size or length extrapolation; the evidence in this paper
instead lands on the expressivity limits of bounded-depth message passing --
the over-squashing line \cite{alon2021bottleneck,topping2022understanding}
and graph rewiring \cite{topping2022understanding,karhadkar2023fosr},
implicit and steady-state GNNs \cite{gu2020implicit}, and graph transformers
\cite{rampasek2022recipe,shirzad2023exphormer} -- since the matched-fit
result says the alternatives cannot represent or attain the target, an
expressivity/attainability story rather than a generalization-preference
one. Section~\ref{sec:program} engages this literature directly as the
competitor set for the resolvent-layer experiment (E3), which is explicitly
an expressivity-separation claim, not an inductive-bias claim. The consumers
that make a differentiable selected inverse useful, deep equilibrium models
\cite{bai2019deep} and delta-rule linear-attention layers
\cite{yang2024parallelizing}, supply the structured symmetric and
non-symmetric solves of Section~\ref{sec:deq} and the appendix.
